---
title: "WriteUp"
output: html_document
---
Dataset from
Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, 
H. Qualitative Activity Recognition of Weight Lifting Exercises. 
Proceedings of 4th International Conference in Cooperation with 
SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013.
Read more: http://groupware.les.inf.puc-rio.br/har#ixzz3uJG8yOEC

We start by loading caret and other potentially useful packages
```{r}
library(caret)
library(corrplot)


```

And then provide remote urls to get the data from and set the working directory 

```{r}

train_url = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
test_url = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv" 

directory = "."
setwd(directory)
trainfile = "train.csv"
testfile = "test.csv"
train_dest = paste(directory, trainfile,sep ="/")
test_dest = paste(directory, testfile,sep ="/")
# if files do not exist then download them from url
if (!file.exists(train_dest) | !file.exists(test_dest)){
method = "curl"  
download.file(train_url,destfile = train_dest,method = method)
download.file(test_url,destfile = test_dest,method = method)
}
```
Read in "raw" datasets

```{r}
train_raw <- read.csv(train_dest)
test_raw <-read.csv(test_dest)
```

 Preprocess datasets

Begin by removing columns with missing values
```{r}
cols = apply(train_raw,2,function(x) {sum(is.na(x))==0})
train_raw = train_raw[,cols]

```
And do the same on the test set
```{r}
test_raw = test_raw[,cols]
```

We now proceed on to removing columns containing "#DIV/0!" as 
it is not a meaningful value and hence will not be useful for prediction
```{r}
cols = apply(train_raw,2,function(x) {length(grep("#D",x))==0} )
train_raw = train_raw[,cols]
test_raw = test_raw[,cols]
```

At this point, after basic cleaning, the two datasets contain 60 columns.
We can further reduce this number by noting that columns containing user names,
time stamps and windows will not be useful for prediction and can thus be 
eliminated

```{r}
train_raw = train_raw[,-grep("user_name|window|timestamp|^X",names(train_raw))]
test_raw = test_raw[,-grep("user_name|window|timestamp|^X",names(test_raw))]
```

Which brings the number of columns down to 53.

We can now perform a slightly more sofisticated pre-processing 
by checking if any pairs of variables is significantly correlated
in which case we can further reduce the dimensionality of the problem

We thus construct the correlation matrix and set its diagonal to 0

```{r}
corrmat = abs(cor(train_raw[,-53]))
diag(corrmat) = 0
```
We visualize the matrix and notice that there is some appreciable
correlation between variables which is a hint that the dimensionality
can be further reduced
```{r}
corrplot(corrmat,method = "color", tl.cex = 0.5)
```

Given the presence on non-negligible correlations we perform
principal component analysis and retain only those
variables contributing to 95% of the variance are retained.
```{r}

  
  preobj = preProcess(train_raw[,-53],method = "pca", thresh = 0.95)
  trainPC = predict(preobj,train_raw[,-53])
  testPC = predict(preobj,test_raw[,-53])
 
  train = trainPC
  test = testPC
 
  
ncol(train)
ncol(test)

```

At the end of this pre-processing phase we have obtained a polished
training and test set. We now leave the test set aside and proceed to 
data slicing, setting a seed for reproducibility.
```{r}
set.seed(20140604)
```

We then split the training dataset in two parts for validation
and correspondingly create two "classe" vectors against which
validate the model

```{r}
cvind = createDataPartition(train_raw$classe,p = 0.7,list = FALSE)
train_cv = train[cvind,]
classe_train_cv = train_raw$classe[cvind]
test_cv = train[-cvind,]
classe_test_cv = train_raw$classe[-cvind]
```

Once the data has been sliced we enter the model training part.
We begin by choosing the parameters for cross validation

```{r}
controls1 = trainControl(method = "cv", number = 5)
```

I chose the number of folds for cross validation to be 5 in order
for it not be too small (hence limit the bias) but still manageable
from a computational point of view. Normally I would have done
few trials trying different values for number.
Also, having the time, it could be interesting to set number = 10
and compare cross validation against repeated cross validation with 
2 repeats and 5 folds both in terms of computational efficiency and
performance of the fitted model.

We now proceed to model training. I choose to use a decision-tree based algorithm
as they
appear to me best suited for multi-variate classification problems. 
I begin by training a decision tree with maximum tree-depth = 30 (which should be
the maximum allowed value)

```{r}
if (!exists("modFitDecTree")){
  modFitDecTree <- train(classe_train_cv ~.,
                   data = train_cv,
                   method="rpart2",
                   trControl = controls1,
                   tuneGrid = data.frame(.maxdepth = 30))
  
  
}
print(modFitDecTree$finalModel)

```

and then a random forest: in order to improve computation speed I decided to 
set mtry, i.e. the variables tried at each split, to be 18.
The reason  is that after PCA all variables, being
linearly independent, are likely to contribute on a split. On the other hand
I wanted to prevent overfitting and gain some computational speed, so
trying  approx 2/3 of the available variables seemed to me a good compromise.

```{r}
if (!exists("modFit1")){
 modFit1 <- train(classe_train_cv ~.,
                 data = train_cv,
                 method="rf",
                 trControl = controls1,
                 prox = TRUE,
                 allowParallel = TRUE, ntrees  = 250, 
                 tuneGrid = data.frame(.mtry = 18))
}
print(modFit1$finalModel)
```

We now check how the fitted models do against
the training set

decision tree
```{r}
predict0dt = predict(modFitDecTree,train_cv)
confusionMatrix(classe_train_cv,predict0dt)
```

random forest
```{r}
predict0rf = predict(modFit1,train_cv)
confusionMatrix(classe_train_cv,predict0rf)
```

From examination of the results, we choose
random forest as prediction algorithm 
due to its improved accuracy. However,
 the fact 
that the training model is fitted with accuracy 
1 can still be interpreted as 
a warning sign of overfitting despite cross validation? 

We now test the performance of the rf  model 
on the part of the training test left out from training
```{r}

predict1 = predict(modFit1,test_cv)

confusionMatrix(classe_test_cv,predict1)
```

Having obtained an estimated accuracy of 97% and out-of-sample error of 3%, 
and taking
into account time/computational limitations, I believe
the algorithm performs sufficiently good to proceed to prediction.
I therefore apply the fitted model to the test set and 
write the answers in separate files

```{r}
predict_on_test = predict(modFit1,test)
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

answers = as.character(predict_on_test)

answers_dest = paste(directory, "prediction",sep ="/")
if (!file.exists(answers_dest)){
  dir.create(answers_dest)
}
setwd(answers_dest)

pml_write_files(answers)
```



